Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2025-01-08 manuscript/manuscript.tex
Updates responding to JY's comments - mostly chart formatting
Updated control on charts to allow for dynamic sizing of x-axis labels and tick marks. Required changes to all the charting functions as well as the two core programs (datacleansing.R and timeline.R). Simplified the basic chart program to accommodate both numeric and categorical data.  Moved revised images to manuscript image file.
Simple edits
Included note on  the attempt to standardize street names, which failed to increase "matching" of cross_street & intersection_street by any meaningful amount.

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile channels in 2009, a
record monhtly high of 348,463 SRs in 
August 2020 (associated with the the COVID-19 
pandemic), the 2021 expansion to include the 
Metropolitan Transit Authority (MTA) NYC subway 
system (the biggest expansion in 311 history), 
and a yearly record of 3.23 million SRs in 2023.

@@

Our investigation utilizes a 311 SR dataset covering  2-years (2022-2023).
This sample was collected on 15 September 2024, and is used 
to conduct detailed analysis in areas such as data 
consistency, data validity, redundant fields, etc.. Instructions 
on downloading this file from the NYC Open Data Portal 
are available in the supplemental material. Here are some 
characteristics of the 311 SR dataset used in this analysis:

\begin{itemize}[left=1.5em]
\item The dataset is approximately 3.7 GB in size and contains 
  over 6.4 million rows. It was exported as a CSV file.

@@

Collectively these six City Agencies handle over 90\% of the total SR 
volume, indicative of the critical role these Agencies play in managing
public concerns, ranging from noise complaints and housing issues to
sanitation and transportation problems. The remaining
10\% of SRs is distributed across 10 additional Agencies. 
This concentration of SRs to the ``big six'' Agencies 
underscores the necessity for optimal data handling and 
curation processes within these  high volume Agencies.

@@

\subsection{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or presented within a dataset. They can make 
it difficult to analyze the data effectively and often require extensive 
data cleaning efforts. Here are three structural issues we encountered.

@@

\subsection{Missing Data by Field}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their
\texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} fields are blank. When sorting fields by
blank (or N/A values) they appear to divide into three groups:

@@

For some analysis efforts, it may be prudent to inquire 
further as to why some fields are mostly or almost 
always blank. Figure~\ref{fig:blank_fields} is 
a graphic depiction of total empty (blank \& N/As) for each 
field illustrating the grouping into the Mostly Empty, 
Partially Empty, and Few/None Empty categories.

@@

However,  some fields proved problematic as to their 
compliance with a domain of allowable values. These include:

@@

\paragraph{Created or Closed at Midnight:}
Another issue involves the unusually large number of 
SRs where the \texttt{created\_date} and \texttt{closed\_date} 
indicate ``created'' or ``closed'' exactly at midnight (00:00:00), 
to the second. Normally, SR creation and closure follows 
the work-day schedule, with most SRs created 
during working hours and fewer at night or in the 
early morning. However, there is a significantly higher 
number of SRs closed exactly at midnight, as well 
as a large number created exactly at midnight. The magnitude 
of this anomaly is well outside the 3$\sigma$ level as shown 
in Figure~\ref{fig:exacthours}

\begin{figure}[tbp]
    \centering
    % First subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs created exactly on the hour}
        \label{fig:busiestcreated}
    \end{subfigure}
    \hfill % Add horizontal space between subfigures
    % Second subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs closed exactly on the hour}
        \label{fig:busiestclosed}
    \end{subfigure}
    % Main caption
    \caption{Comparison of 311 SRs created \& closed by hour-of-the-day during 2022--2023.}
    \label{fig:exacthours}
\end{figure}
\jy{Figure 4 and 5: Increase the fonts of the axis ticks. This might
  be caused by the size of the pdf figure being too big relative to
  the size shown in the compiled manuscript. So, keep the target size
  in mind may help, avoiding overstretching. In addition, the x-axis
  ticks could be made upright.}

These unusual patterns of SR creation and closure exactly at midnight 
suggest the presence of a bulk create/close software process, perhaps one
that automatically assigns time stamps of midnight (00:00:00) to 
large batches of SRs prior to sending the data to the 311 
data warehouse. This behavior distorts the calculated duration of these 
SRs as well as overall system usage by providing inaccurate 
dates. Further analysis of the distribution by Agency shows that over 92\% of 
the ``closed-exactly-at-midnight'' SRs come from just two 
Agencies: DOB (69\%) and DSNY  (23\%).

@@

\subsection{Accuracy and Precision:}
\label{sec:precision}
An question of precision vs. accuracy arises with the \texttt{latitude} 
and \texttt{longitude} fields. as well as with the duplicative 
\texttt{location} field. These fields are expressed as
a 14-decimal number, e.g. \texttt{latitude} 40.86769186022511. 
Given that 1 degree of latitude at the equator is equal to 111.04 
kilometers, the ``1'' at the end of the decimal number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). (For 
reference a DNA molecule is approximately 2 nanometers in width.) The 
representation of the \texttt{latitude} and \texttt{longitude} fields
appears to be a classic case of extreme precision, but limited 
accuracy. All the \texttt{latitude}, \texttt{longitude}, and \texttt{location}
fields display this characteristic.

\subsection{Redundant Fields}
\label{sec:duplicates}
During this analysis, several redundant fields were identified and should 
be examined for possible consolidation.

@@

\subsection{Reducing File Size}
\label{sec:filesize}
By removing duplicate and ``near-duplicate'' fields, it is possible to 
shrink the file size by 39.5\%, which for this large dataset equates to 
a reduction of 1.4GB, a significant amount. A smaller file size 
means faster downloads, less storage impact, and often 
simpler data analysis efforts. Below is a list of duplicate 
and near-duplicate fields which we believe could be
removed from the dataset with minimal data loss.

@@

    \item \texttt{park\_borough}: This field is a 100\% match with 
    the \texttt{borough} field and can be deleted without data loss.

@@

    \item \texttt{cross\_street\_1/2} and \texttt{intersection\_street\_1/2}: 
    These two street pairs both exhibit an 88\% match. While we cannot 
    confirm the accuracy of either of these street pairs, we recommend deleting the 
    \texttt{intersection\_street\_1/2} fields while acknowledging 
    the concomitant data loss.
\end{itemize}

@@

\paragraph{Field Encoding:} Further reduction is possible 
by encoding selected categorical variables 
Using \texttt{complaint\_type} as an example,
instead of repeatedly storing the full text of each \texttt{complaint\_type}, 
one could use a numeric representation. For example, 
``NOISE - STREET/SIDEWALK'' could be represented by a 
numeric code (e.g. ``126''). For this dataset the 
``NOISE - STREET/SIDEWALK'' \texttt{complaint\_type} occurs 
over 300,000 times. Such encoding not only reduces file 
size, but would likely improve processing efficiency by using numeric
values instead of text strings. Moreover, standardizing 
categorical inputs prevents the inclusion of misspellings 
or unrecognized/invalid entries, ensuring cleaner, more reliable 
data. This approach is particularly useful in large datasets, where 
frequently repeated text-based categorical variables can 
significantly increase storage requirements.

One promising alternative to reducing data storage requirements 
for this large 311 SR dataset (3.7GB) would be employment of 
the cross-platform Apache Arrow package. Apache Arrow (arrow.apache.org) 
is an innovative, cross-language development platform designed for 
in-memory data. It defines a standard for representing columnar data, 
enabling efficient analytics and real-time data processing. As such,
it offers dramatic improvements in file storage size, as well as 
efficient analysis and cross-platform support for R, Python, 
and Julia programming languages. Remarkable storage size 
reduction (within several orders of magnitude) is possible \citep{bates2024csv}.

@@

The 311 SR Data Dictionary would greatly benefit from an update. We
found several discrepancies between the Data Dictionary and the actual
data, which could potentially mislead analytic efforts. Here are some
examples:

@@

\section{Government Open Data Curation Recommendations}
\label{sec:recommendations}
Building upon the insights gained from our case study of the 
NYC 311 SR data, we propose a set of data 
curation principles tailored for government released open datasets. 
These principles are designed to address the unique challenges 
and requirements observed in the curation of such datasets, 
ensuring that they remain reliable, consistent, and useful for 
public services, research, and other applications.

\subsection{Principles}
\paragraph{Principle 1: Establishing Cross-Orginazational Consistency}
A key challenge in government released datasets is the 
harmonization of data fields across various Agencies and 
maintaining this consistency over time. To address this, agencies should adopt 
standardized naming conventions, value domains, usages, and 
formats across all datasets. Regular audits and data harmonization 
processes should be implemented to ensure consistency 
across long-term datasets, particularly as agency structures 
or reporting standards evolve. For instance, ensuring that 
historical datasets reflect the same field naming conventions 
over a span of years can prevent inconsistent analyses. 
These standards should be promulgated by the Open Data 
governing authority, following agreements by the relevant parties.

@@

\begin{itemize}[left=1.5em]
    \item Certain date fields should be promulgated by the software
    system versus manual entry. Critical dates should be populated
    based upon an action in the system, such as a change in status.

    \item Select fields should be subjected to automated checks 
    that prevent logical errors, such as closing and opening.

@@

\paragraph{Principle 4: Updating Data Dictionaries}
Clear and accurate data documentation is essential for the effective use 
of government released datasets. Our study found several 
instances where the Data Dictionary did not accurately reflect the 
actual data. Domains of legal/allowable values should be referenced
in the Data Dictionary. Regular reviews and updates should 
occur whenever there are changes in data structure, data usage, or 
content. This practice enhances the ability of external users, 
including researchers and developers, to understand and work 
effectively with the data.

@@

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset(s) structure is maintained over time. Their expertise is also 
critical in developing protocols for validating incoming data, correcting 
errors in real-time, and ensuring consistency across multiple data sources. 
Feedback from statisticians and data scientists, such as the issues reported 
in this paper, benefit from a clear and accessible channel to the 311 SR data 
team and other relevant City Agencies. Establishing such feedback loops 
and other opportunities for public engagement, ensures that 
recommendations for improving data curation and validation 
are not only heard and discussed, but implemented. This feedback 
mechanism would also allow ongoing dialogue between 
data users and City officials, leading to more dynamic 
and responsive data management practices.

A particularly valuable avenue for engagement is through events such as NYC 
Open Data Week, which fosters collaboration and innovation across sectors. 
Our involvement in this project stemmed from such an event, where we were 
inspired by the potential of open data to enhance public services. Governments 
should continue supporting these types of initiatives, which not only promote 
data literacy, but encourage community-driven improvements to datasets. 
By leveraging the expertise of data scientists, statisticians, stakeholders, 
and civic organizations, governments can ensure that their open datasets 
remain accurate, transparent, and useful for a wide range of applications.

>>>>

The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile channels in 2009, a
record monthly high of 348,463 SRs in 
August 2020 (associated with the the COVID-19 
pandemic), the 2021 expansion to include the 
Metropolitan Transit Authority (MTA) NYC subway 
system (the biggest expansion in 311 history), 
and a yearly record of 3.23 million SRs in 2023.

@@

Our investigation uses a 311 SR dataset covering  2-years (2022-2023), 
and was collected on 15 September 2024. It is to conduct 
analysis in areas such as data consistency, data validity, 
redundant fields, etc.. Instructions on downloading this file are 
available in the supplemental material. Characteristics of the 311 SR dataset include:

\begin{itemize}[left=1.5em]
\item The dataset is 3.7 GB in size and contains 
  over 6.4 million rows. It was exported as a CSV file.

@@

Collectively these six City Agencies handle over 90\% of the total SR 
volume, indicative of the critical role they play in managing
public concerns; ranging from noise complaints and housing issues to
sanitation and transportation issues. The remaining
10\% of SRs are distributed across 10 additional Agencies. 
This concentration of SRs to these ``big six'' Agencies 
underscores the necessity for optimal data curation processes 
within these  high volume Agencies.

@@

\subsection{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or presented within a dataset. Structural issues can cause difficulty in 
 analyzing the data effectively, often necessitating extensive 
data cleaning. We encountered three structural issues.

@@

\subsection{Missing Data by Field}
\label{sec:blanks}
Understanding the absence of data by field greatly aids when 
conducting analysis. For example, determining if SRs were closed 
before or after their \texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} fields are blank, and are only used by one Agency, DSNY. Blank 
(or N/A values) they appear to divide into three groups:

@@

For some analysis efforts, it may be prudent to inquire 
 as to why some fields are mostly/almost always blank. 
Figure~\ref{fig:blank_fields} is a graphic depiction of the \%
empty (blank \& N/As) for each field. It illustrates the grouping into 
Mostly Empty, Partially Empty, and Few/None Empty segments.

@@

However, some fields lacked compliance
to a domain of allowable values, including:

@@

\paragraph{Created or Closed at Midnight:}
Another issue involves the unusually large number of 
SRs where the \texttt{created\_date} and \texttt{closed\_date} 
indicate action exactly at midnight (00:00:00), 
to the second. Normally, SR creation and closure follows 
the work-day schedule, with most SRs created 
during working hours and fewer at night or in the 
early morning. However, there is a significantly higher 
number of SRs closed exactly at midnight, as well 
as a large number created exactly at midnight. The magnitude 
of this anomaly is well outside the 3$\sigma$ level as shown 
in Figure~\ref{fig:exacthours}

\begin{figure}[tbp]
    \centering
    % First subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_created_on_the_hour.pdf}
        \caption{SRs created exactly on the hour}
        \label{fig:busiestcreated}
    \end{subfigure}
    \hfill % Add horizontal space between subfigures
    % Second subfigure
    \begin{subfigure}[t]{0.495\textwidth} % Adjust width to fit side by side
        \centering
        \includegraphics[width=\textwidth]{2-year-trend_SRs_closed_on_the_hour.pdf}
        \caption{SRs closed exactly on the hour}
        \label{fig:busiestclosed}
    \end{subfigure}
    % Main caption
    \caption{Comparison of 311 SRs created \& closed by hour-of-the-day during 2022--2023.}
    \label{fig:exacthours}
\end{figure}

These unusual patterns of SR creation/closure at midnight 
suggest the presence of a bulk create/close software process, perhaps one
that automatically assigns time stamps of midnight (00:00:00) to 
 batches of SRs prior to sending the data to the 311 
database. This behavior distorts the calculated duration of these 
SRs as well as overall system usage by providing inaccurate 
dates. Further analysis of the distribution by Agency shows that over 92\% of 
the ``closed-exactly-at-midnight'' SRs come from just two 
Agencies: DOB (69\%) and DSNY  (23\%).

@@

\subsection{Accuracy and Precision:}
\label{sec:precision}
The question of precision vs. accuracy arises with the \texttt{latitude} 
and \texttt{longitude} fields. as well as with the duplicative 
\texttt{location} field. These fields are expressed as
a 14-decimal number, e.g. \texttt{latitude} 40.86769186022511. 
Given that 1 degree of latitude at the equator is equal to 111.04 
kilometers, the ``1'' at the end of that decimal number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). (For 
reference a DNA molecule is approximately 2 nanometers in width.) This 
representation appears to be a classic case of extreme precision, albeit with limited 
accuracy. All the \texttt{latitude}, \texttt{longitude}, and \texttt{location}
fields display this characteristic.

\subsection{Redundant Fields}
\label{sec:duplicates}
This analysis revealed several redundant fields, indicting a potential need 
for consolidation.

@@

\begin{quote}
\textit{Note: The authors attempted extensive standardization of street names  
in an attempt to match what were obvious identical streets, deemed unequal 
due only to spelling and naming conventions. Those efforts were intended to match, for example, 
"KANE PL" and "KANE PLACE" as well as "FLATLANDS 7 ST" and "FLATLANDS 7 STREET". However, these 
efforts were  abandoned after they increased street name matches by less than 1\%.)}
\end{quote}

\subsection{Reducing File Size}
\label{sec:filesize}
By removing duplicate and near-duplicate fields, one can reduce file size by 39.5\%, 
equating to a reduction of 1.4GB, a significant amount. A smaller file size 
means faster downloads, reduced storage impact, and often simpler data analysis. 
Below is a list of duplicate and near-duplicate fields which we believe could be
removed from the dataset with minimal data loss.

@@

    \item \texttt{park\_borough}: This field is a 100\% match with 
    the \texttt{borough} field. It can be deleted without data loss.

@@

    \item \texttt{cross\_street\_1/2} and \texttt{intersection\_street\_1/2}: 
    These two street pairs exhibit an 88\% match. Although we cannot 
    confirm the accuracy of either of these street pairs, we recommend deleting the 
    \texttt{intersection\_street\_1/2} fields while acknowledging 
    the concomitant data loss.
\end{itemize}

@@

\paragraph{Field Encoding:} Further size reduction is possible 
by encoding selected categorical variables 
Using \texttt{complaint\_type} as an example,
instead of repeatedly storing the full text of each \texttt{complaint\_type}, 
one could use a numeric representation. For example, 
``NOISE - STREET/SIDEWALK'' could be represented by a 
numeric code (e.g. ``126''). For this dataset the 
``NOISE - STREET/SIDEWALK'' \texttt{complaint\_type} occurs 
over 300,000 times. Encoding not only reduces file 
size, but would likely improve processing efficiency by using numeric
values instead of text strings. Moreover, standardized 
categorical inputs often prevents misspellings 
and invalid entries, ensuring cleaner, more reliable 
data. This approach is particularly useful in large datasets, where 
frequently repeated text-based categorical variables significantly 
increases storage requirements.

One promising alternative to reducing data storage  
for this large 311 SR dataset (3.7GB) would be employment of 
the cross-platform Apache Arrow package. Apache Arrow (arrow.apache.org) 
is an innovative, cross-language development platform designed for 
in-memory data. It defines a standard for representing columnar data, 
enabling efficient analytics and real-time data processing. As such,
it offers dramatic improvements in file storage size, as well as 
efficient analysis and cross-platform support for R, Python, 
and Julia programming languages. Remarkable storage size 
reduction (within several orders of magnitude) is possible \citep{bates2024csv}.

@@

The 311 SR Data Dictionary would greatly benefit from an update. We
found several discrepancies between the Data Dictionary and the actual
data. Such errors could potentially mislead analytic efforts. Examples include:

@@

\section{Government Open Data Curation Recommendations}
\label{sec:recommendations}
Building upon  insights from our case study of the 
NYC 311 SR data, we propose a set of data 
curation principles tailored for government open datasets. 
These principles are designed to address the unique challenges 
 observed in the curation of such datasets, 
ensuring they remain reliable, consistent, and useful for 
public services, research, and other applications.

\subsection{Principles}
\paragraph{Principle 1: Establishing Cross-Organizational Consistency}
A key challenge in government released datasets is the 
harmonization of data fields across various Agencies, and 
maintaining that consistency over time. To address this, Agencies should adopt 
standardized naming conventions, value domains, usages, and 
formats across all datasets. Regular audits and data harmonization 
processes should be implemented to ensure consistency 
across long-term datasets, particularly as Agency structures 
and reporting standards evolve. For instance, ensuring that 
historical datasets reflect the same field naming conventions 
over a span of years can prevent inconsistent analyses. 
These standards should be promulgated by the Open Data 
governing authority, following agreements by the relevant parties.

@@

\begin{itemize}[left=1.5em]
    \item Certain date fields should be promulgated by the application software
     versus manual entry. Dates should be populated
    based upon an action in the system, such as a change in status.

    \item Select fields should be subjected to automated checks 
    that prevent logical errors, such as a closing date occurring before the creation date.

@@

\paragraph{Principle 4: Updating Data Dictionaries}
Clear and accurate data documentation is essential for the effective use 
of government datasets. Our study found several 
instances where the Data Dictionary did not accurately reflect the 
actual data. Domains of legal/allowable values should be referenced
in the Data Dictionary. Regular reviews and updates should 
occur whenever there are changes in data structure, data usage, or 
content. This practice enhances the ability of external users, 
including researchers and developers, to work 
effectively with the data.

@@

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset(s) structure is maintained over time. Their expertise is also 
critical in developing protocols for validating incoming data, correcting 
errors in real-time, and ensuring consistency across multiple data sources. 
Feedback, such as the issues reported 
in this paper, benefit from a clear and accessible channel to the 311 SR data 
team and other relevant City Agencies. Establishing such feedback loops 
and other opportunities for public engagement, ensures that 
recommendations for improving data curation and validation 
are not only heard and discussed, but implemented. This feedback 
mechanism would also allow ongoing dialogue between 
data users and City officials, leading to more dynamic 
and responsive data management practices.

An especially valuable engagement avenue is through events such as NYC 
Open Data Week, which foster collaboration and innovation across sectors. 
Our involvement in this project stemmed from such an event, where we were 
inspired by the potential of open data to enhance public services. Governments 
should continue these types of initiatives, which not only promote 
data literacy, but encourage community-driven improvements to datasets. 
By leveraging the expertise of data scientists, statisticians, stakeholders, 
and civic organizations, governments can ensure their open datasets 
remain accurate, transparent, and useful for a wide range of applications.
